---
title: "Probability, Part II"
author: "<small><br>Stephen Vaisey<br>Duke University</small>"
date: "<span style = 'font-size: 65%;'>Last update: `r Sys.Date()`<br><br>`r icon::fa_link()` <a href='stephenvaisey.com'><font color='F3F2F1'>stephenvaisey.com</font></a><br>`r icon::fa_twitter()` <a href='http://twitter.com/vaiseys'><font color='F3F2F1'>@vaiseys</font></a><br>`r icon::fa_github()` <a href='https://github.com/vaiseys'><font color='F3F2F1'>vaiseys</font></a></span> "
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, addons.css]
    self_contained: false
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
---

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.asp = 0.618,
  fig.retina = 3,
  fig.width = 6,
  message = FALSE,
  warning = FALSE,
  dev = "svg",
  out.width = "80%"
)
options(knitr.table.format = "html")
options(knitr.kable.NA = '   ')

library(here)
library(knitr)
library(tidyverse)
library(broom)
library(DiagrammeR)
theme_set(theme_minimal(base_family = "Lato"))

```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#00539B" ,
  white_color = "#F3F2F1" ,
  black_color = "#262626" ,
  header_h1_font_size = "110px",
  header_h2_font_size = "60px",
  header_h3_font_size = "45px",
  text_font_size = "28px" ,
  code_font_size = ".65em" ,
  header_color = "#C84E00" ,
  text_slide_number_font_size = "0.6em" ,
  code_highlight_color = "rgba(161,183,13,0.5)" ,
  link_color = "#993399" ,
  text_font_google = google_font("Lato", "400", "400i", "700") ,
  code_font_google = google_font("Roboto Mono", "400") ,
  extra_css = list(
    ".remark-slide-content" = list("padding-top" = "60px") ,
    "h1" = list("margin" = "0 0 20px 0") ,
    "h2" = list("margin" = "0 0 20px 0") ,
    "h3" = list("margin" = "0 0 20px 0") ,
    "th, td" = list( "padding" = "10px" ) )
)

```

class: center, middle

# Ask two questions

---

class: center, middle

# Assessment time

---

class: center,middle,inverse

# Probability

---

## Permutations

.pull-left[

$$_nP_k=\frac{n!}{n-k!}$$

]

.pull-right[

There are 5 groups to present in class. How many different presentation orders 
are possible? That is, how many ways can we arrange it?

$5\times 4 \times 3 \times 2 \times 1 = 120$

]

???

Ordering matters

5 permutations of 5 = 5!/0!

"Out of a set of 5, how many different ways can we order 5?"

---

## Combinations

.pull-left[

$$_nC_k = {n \choose k} = \frac{_nP_k}{k!} = \frac{n!}{k!(n-k)!} $$

Pronounced "_n_ choose _k_"

]

.pull-right[

If there are six faculty available, how many different committees of 3 can
be formed?

$$\frac{6!}{3!3!} = 20$$

From a department of 20 professors, how many different committees or 5 can be 
formed?

```{r, echo = TRUE}
choose(20, 5)

```
]

???

Ordering doesn't matter

(6x5x4x3x2)/((3x2)(3x2)) = (6x5x4)/(3x2) = 20

---

## Combinations and probablity

$$_nC_k = {n \choose k} = \frac{_nP_k}{k!} = \frac{n!}{k!(n-k)!} $$

Let’s say 3 professors are male and 3 are female. What’s the probability that a 
committee of 3 with zero men will be formed?

How many ways are there to choose zero men out of 3? How many ways to choose 3 
women out of three? How many total ways are there for **anything** to happen?

--

$$\frac{{_3C_0} \times {_3C_3}}{_6C_3} = \frac{1 \times 1}{20} = .05 $$
