---
title: "Prediction: Part II"
author: "<br>Stephen Vaisey<br>Duke University"
date: "<small><small>Fall 2019<br><br>`r icon::fa_link()` <a href='stephenvaisey.com'><font color='F3F2F1'>stephenvaisey.com</font></a><br>`r icon::fa_twitter()` <a href='http://twitter.com/vaiseys'><font color='F3F2F1'>@vaiseys</font></a><br>`r icon::fa_github()` <a href='https://github.com/vaiseys'><font color='F3F2F1'>@vaiseys</font></a></small></small> "
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["left", "middle", "inverse"]
---

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, fig.align = 'center', fig.width = 6, fig.asp = .618 , out.width = "80%", fig.retina = 3 , dev = 'svg')

options(knitr.table.format = "html")

library(here)
library(tidyverse)
theme_set(theme_minimal())

```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#00539B" ,
  white_color = "#F3F2F1" ,
  black_color = "#262626" ,
  header_h1_font_size = "75px",
  header_h2_font_size = "60px",
  header_h3_font_size = "45px",
  text_font_size = "30px" ,
  code_font_size = ".9em" ,
  code_font_family = "Input" ,
  header_color = "#C84E00" ,
  text_slide_number_font_size = "0.6em" ,
  code_highlight_color = "rgba(161,183,13,0.5)" ,
  link_color = "#993399" ,
  extra_css = list(
    ".remark-slide-content" = list("padding-top" = "50px") ,
    "h1" = list("margin" = "0 0 20px 0") ,
    "h2" = list("margin" = "0 0 20px 0") ,
    "h3" = list("margin" = "0 0 20px 0") )
  )

```


# Homework debrief

Ask **two questions** about the concepts

---

# Today's agenda

1. Review of correlation and regression

2. More on linear regression

---
class: middle, inverse, center

# 1. Correlation and regression

---

# Correlation

- On average, how do two variables move together

- Positive (negative) correlation: When x is larger than its mean, y is likely (unlikely) to be larger than its mean

- Positive (negative) correlation: data cloud slopes up (down)

- High correlation: data cluster tightly around a line

---

# Visualizing different correlations

```{r diffcorrs}
# make a function that makes fake data with a given correlation
mkdata <- function(n, r) {
  result <- 
    MASS::mvrnorm(n, 
                  mu = c(0,0), 
                  Sigma = matrix(c(1,r, r,1) , 
                            nrow = 2 ,
                            dimnames = list(c("y","x"))) ) %>% 
    as_tibble() %>% 
    mutate( corr = r )
  return( result )
}

# for reproducibility
set.seed(0903)

# make four different versions
cdata <- bind_rows( mkdata( 1000 ,  0 ) ,
                    mkdata( 1000 , .3 ) ,
                    mkdata( 1000 , .6 ) ,
                    mkdata( 1000 , .9 ))

# plot the result
ggplot( cdata , aes( x = x, y = y )) +
  geom_point( alpha = .3 ) +
  facet_wrap( ~factor(corr) ) 


```

---

# Linear regression model

$$Y = \alpha + \beta X + \varepsilon$$

<small>

| Symbol|  Meaning |
|:------|---------:|
| $Y$ | outcome variable |
| $X$ | predictor variable |
| $\alpha$, $\beta$ | parameters of the model |
| $\varepsilon$ | unobserved random error |
| | |
| $\alpha+\beta X$ | expected mean of $Y$ given a value of $X$  |
| $\alpha$ | expected value of $Y$ when $X=0$ |
| $\beta$ | expected change in $Y$ given 1-unit increase in $X$ |

</small>

---

# 
# Some algebra for least squares

### Estimated coefficients in mathematical expressions

$$\hat{\beta} = \frac{\sum(Y_i - \bar{Y}) (X_i - \bar{X})} {\sum(X_i - \bar{X})}$$

$$\hat{\alpha} = \bar{Y} - \hat{\beta}\bar{X} $$

--

### Relationship between slope and correlation

$$\hat{\beta} = corr(Y,X) \times \frac{S_y}{S_x} $$

---

# Why _this_ line?

### Estimate the model parameters from the data
$(\hat{\alpha},\hat{\beta})$ are estimated coefficients <br>
$\hat{Y} = \hat{\alpha}+\hat{\beta}X$ are fitted/predicted values <br>
$\hat{\varepsilon} = Y - \hat{Y}$ are estimated residuals

### We obtain these estimates via least squares

That is, minimizing $SSR = \sum\hat{\varepsilon_i}^2$ or $RMSE= \sqrt{\frac{SSR}{n}}$

---

# How well does the model fit?

### In other words, how well does it predict the outcome?

$~$

$$R^2 = 1 - \frac{SSR}{TSS} = 1-\frac{\sum \hat{\varepsilon}_i^2}{\sum(Y_i - \bar{Y})}$$

.footnote[This is the correlation coefficient (_r_) squared. Here's that's about
.76. You could say "differences in life expectancy account for 76% of the 
variance in fertility." But remember, this is **descriptive**, not **causal**.]

---

# Remember!

### Regression and correlation assume linearity

---

# Don't do this!

```{r notlinear}
data("gapminder", package = "dslabs")
gm2011 <- gapminder %>% 
  filter( year == 2011 ) %>%
  mutate( gdpcap = (gdp/population) / 1000 ) %>% 
  drop_na()

ggplot( gm2011 , aes( x = gdpcap , y = fertility ) ) +
  geom_point( alpha = .3 ) +
  geom_line( stat = "smooth" , 
             method = "lm" ,
             alpha = .5 , 
             color = "#262626" ) +
  labs( title = "Fertility by GDP in 185 countries, 2011" ,
        x = "GDP per capita in 1000s of Euros" ,
        y = "Fertility" )

```

---

# This is a bit better

```{r loglinear}
gm2011 <- gm2011 %>% 
  mutate( log10gdpcap = log10(gdp/population) )

ggplot( gm2011 , aes ( x = log10gdpcap , y = fertility ) ) +
    geom_point( alpha = .3 ) +
  geom_line( stat = "smooth" , 
             method = "lm" ,
             alpha = .5 , 
             color = "#262626" ) +
  labs( title = "Fertility by GDP in 185 countries, 2011" ,
        x = "log(10) GDP per capita (Euros)" ,
        y = "Fertility" )

```

---
class: inverse, middle, center

# 2. Regression and causality

---

# Dictator game

- You get 10 raffle tickets to win a prize. You are paired with another 
respondent who gets no tickets. You can share as many or as few tickets as you want. How many do you want to share?

--

- Experimental condition: is the raffle worth 50 dollars (_v500_ = 0) or 500 
dollars (_v500_ = 1)?

--

- Research question: are people less likely to share when the raffle is worth 
more?

---

# The model

$~$

$$tickets_i = \alpha + \beta v500_i + \varepsilon_i$$

$~$

Are we allowed to give this $\beta$ a causal interpretation or not? Why or why not?

---

# The distribution

```{r}
# d <- haven::read_dta(here( "data" , "trolley-dictator.dta" )) %>% 
#   haven::zap_labels()
# ggplot( d , aes( x =   ))


```

